---
title: "ARIMAX time series regression"
author: "Victoria Zaitceva"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(tseries)
library(forecast)


data <- read.table("data/data_summarized_by_month_filled_hosp_MR.tsv", header = TRUE, sep = "\t")
```


data_summarized_by_month_filled_hosp_MR.tsv - nothing is significant
data/data_summarized_by_month_filled_hosp.tsv - covid significant, thromb_perc borderline
data/data_summarized_by_month_filled.tsv -covid borderline

data/data_summarized_by_month_filled_all_variables.tsv - ishemic_stroke_perc and covid significant


```{r}
data_prepared <- data %>%
  mutate(
    year_month = anytime::anydate(paste0(year, "-", sprintf("%02d", month), "-01"))
  ) %>%
  group_by(year_month) %>%
  summarise(
    lethal_outcomes_perc = sum(Летальные_Исходы, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    ishemic_stroke_perc = sum(ИИ_Ишемический_Инсульт, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    gem_stroke_perc = sum(ГИ_Геморрагический_Инсульт, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    age60_perc = sum(Старше_60_лет, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    severeNIHSSS_perc = sum(NIHSS_Больше_21, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    ii_therapeutic_window_perc = sum(ИИ_Терапевтическое_Окно, na.rm = TRUE) / sum(ИИ_Ишемический_Инсульт, na.rm = TRUE) * 100,
    gi_therapeutic_window_perc = sum(ГИ_Терапевтическое_Окно, na.rm = TRUE) / sum(ГИ_Геморрагический_Инсульт, na.rm = TRUE) * 100,
    total_therapeutic_window_perc = sum(Терапевтическое_Окно_4_5ч, na.rm = TRUE) / sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE) * 100,
    thromb_perc = sum(Тромболизис, na.rm = TRUE) / sum(ИИ_Ишемический_Инсульт, na.rm = TRUE) * 100,
    total_treated = sum(Пролечены_с_ОНМК_Всего, na.rm = TRUE),
    total_load = sum(Суммарная_Загрузка_ОНМК, na.rm = TRUE),
    #total_not_treated = total_load - total_treated, # try instead of total load - didn't help
    .groups = "drop"
  )
```

COVID variable

```{r}
# 01 January 2020 - December 2021 - COVID period

data_prepared$covid <- ifelse(data_prepared$year_month >= as.Date("2020-03-01") & data_prepared$year_month <= as.Date("2021-12-31"), 1, 0)
```



```{r}
str(data_prepared)
```

# ARIMAX

ARIMA with exogenous regressors (ARIMAX)

## Data preparation

```{r}
# Convert the dependent variable (Deaths) into a time series object
deaths_ts <- ts(data_prepared$lethal_outcomes_perc, start = c(2017, 1), frequency = 12)  # Monthly data

```


```{r}
# Save the independent variables
kpis <- data_prepared[, c("ishemic_stroke_perc", "gem_stroke_perc", "age60_perc", "severeNIHSSS_perc", 
                          "total_therapeutic_window_perc", "thromb_perc", "total_treated" ,"total_load")]
kpis <- as.data.frame(lapply(kpis, as.numeric))
# Convert KPIs to a numeric matrix
kpis_matrix <- as.matrix(kpis)

```


## Stationarity test

```{r}

deaths_ts <- na.omit(deaths_ts)

# Test for stationarity 
adf.test(deaths_ts)
# p-value = 0.2 - data is not stationary

# If p-value > 0.05, differencing is needed
# Apply differencing to make the series stationary
deaths_diff <- diff(deaths_ts)
adf.test(deaths_diff)
# p-value = 0.01 - data is stationary, we can use ARIMA model


```
## Model estimation

```{r}

# Ensure the lengths of deaths_ts and kpis_matrix match
if (length(deaths_ts) != nrow(kpis_matrix)) {
  kpis_matrix <- kpis_matrix[-1, ]  # Align KPIs with the differenced time series
}

arimax_model <- auto.arima(deaths_ts, d = 1, xreg = kpis_matrix)

summary(arimax_model)

```




ARIMA(0,1,1):

p=0: No autoregressive terms are included.
d=1: First differencing is applied to make the series stationary.
q=1: 1 moving average terms account for past forecast errors.


t-statistic = estimate / standard error
if smaller than |1.96|, then the variable is not significant

Significant variables:
Without COVID variable:
ishemic_stroke_perc	-0.2926	/0.0852	= -3.43	Yes
total_treated	0.0047 /	0.0023 =	2.04	Yes
total_load	-0.0031 /	0.0016 = -1.94	No, but close to the threshold	



With COVID variable:
ishemic_stroke_perc	-0.2739	/0.0838	= -3.27	Yes
COVID = 2.9368 / 1.0725 = 2.74	Yes

The rest of the variables are not significant.


Impact of covid:

The coefficient for covid is 2.9368, meaning deaths increase by approximately 2.94 units during COVID-19 periods, holding other factors constant.
Significant t-value (2.74) suggests that the COVID-19 period has a measurable and meaningful impact on deaths.

ishemic_stroke_perc (−0.2739):
Significant negative relationship: Higher ischemic stroke percentage correlates with fewer deaths.

ACF1 in the output: 0.0778
This value measures the lag-1 autocorrelation of residuals.
A value close to 0 indicates minimal autocorrelation.
For practical purposes, ∣ACF1∣<0.2 is typically considered low enough to conclude no substantial autocorrelation.

ma1 (−0.5998):
Significant moving average term, indicating past forecast errors are critical for prediction.

## Model diagnostics

```{r}
# Residual diagnostics
checkresiduals(arimax_model)

```

1. Residuals should fluctuate randomly around zero. The residuals fluctuate around zero, which is good and suggests the model is not biased. However, there appears to be a spike (possibly due to outliers or unusual events) around 2020, likely linked to the onset of COVID-19. No strong trends are visible, which supports the assumption that differencing has effectively removed non-stationarity.

2. Autocorrelation Function (ACF) plot - Correlation of residuals with their lags
No significant spikes (beyond the dashed line) indicate minimal autocorrelation.

3. Residual Histogram
While normality isn't strictly required for ARIMA, the approximate normality of residuals supports the validity of the model.

4. Ljung-Box Test
Ljung-Box Test  tests the null hypothesis that residuals are independently distributed.
Slight skewness in the residual distribution could indicate some outliers 

```{r}
Box.test(residuals(arimax_model), lag = 10, type = "Ljung-Box")

```
p>0.05: Fail to reject the null, indicating no significant autocorrelation.


### Refit the model only with significant variables with COVID


```{r}
selected_kpis <- data_prepared[, c("ishemic_stroke_perc",  "covid")]


selected_kpis_matrix <- as.matrix(selected_kpis)


# Ensure the lengths of deaths_ts and kpis_matrix match
if (length(deaths_ts) != nrow(selected_kpis_matrix)) {
  selected_kpis_matrix <- selected_kpis_matrix[-1, ]  # Align KPIs with the differenced time series
}



simplified_model_covid <- auto.arima(deaths_ts, d = 1, xreg = selected_kpis_matrix)


summary(simplified_model_covid)
```


ar1 (t=2.85):
Significant positive autoregressive term, showing that past values influence current values.

ma1 (t=−12.56):
Strongly significant moving average term, indicating the importance of past forecast errors in explaining the series.

ishemic_stroke_perc (t=−4.32):
Significant negative relationship: Higher ischemic stroke percentages correlate with fewer deaths.

covid (t=3.37):
Significant positive relationship: During COVID-19 periods (covid=1), deaths increase significantly by approximately 3.17 units, holding other factors constant.


Training Set Error Measures
ME	-0.0250	Minimal bias in the predictions.
RMSE	2.2762	Improved accuracy compared to earlier models.
MAE	1.5686	Lowest among all tested models, indicating better fit.
MAPE	8.26%	Excellent predictive accuracy (below 10%).
ACF1	
0.0066	Residual autocorrelation is near zero, indicating no significant unexplained patterns.

```{r}
# Residual diagnostics
checkresiduals(simplified_model_covid)
```




### Refit the model only with significant variables without COVID


```{r}

selected_kpis <- data_prepared[, c("ishemic_stroke_perc", "total_treated", "total_load")]


selected_kpis_matrix <- as.matrix(selected_kpis)


# Ensure the lengths of deaths_ts and kpis_matrix match
if (length(deaths_ts) != nrow(selected_kpis_matrix)) {
  selected_kpis_matrix <- selected_kpis_matrix[-1, ]  # Align KPIs with the differenced time series
}



simplified_model <- auto.arima(deaths_ts, d = 1, xreg = selected_kpis_matrix)


summary(simplified_model)

```

ishemic_stroke_perc  = -0.2261 / 0.0620 = -3.65	Yes
total_treated = 0.0035 / 0.0022 =  1.59	No
total_load = -0.0024 / 0.0015 = -1.61	No




# Forecasting

Теперь ishemic_stroke_perc положительно влияет на смертность, а не отрицательно, как раньше......


```{r}

training_data <- data_prepared[data_prepared$year_month < "2024-01-01", ]
test_data <- data_prepared[data_prepared$year_month >= "2024-01-01", ]

deaths_ts_train <- ts(training_data$lethal_outcomes_perc, start = c(2017, 1), frequency = 12) 

selected_kpis_train <- training_data[, c("ishemic_stroke_perc")]
selected_kpis_matrix_train <- as.matrix(selected_kpis_train)

```

training data 

```{r}

simplified_model <- auto.arima(deaths_ts_train, d = 1, xreg = selected_kpis_matrix_train)

summary(simplified_model)

```


```{r}

selected_kpis_test <- test_data[, c("ishemic_stroke_perc")]
selected_kpis_matrix_test <- as.matrix(selected_kpis_test)


forecast_2024 <- forecast(simplified_model, xreg = selected_kpis_matrix_test, h = nrow(selected_kpis_test))


print(forecast_2024)
autoplot(forecast_2024)

```

